---
title: "Reproducible Research Project1"
author: "Javier A Diaz"
date: "26 de junio de 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is a project of Reproducible Research with the objective of data analysis of information of steps done by a volunteer  during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day. The data was provided from Activity monitoring data set. 

## Data adquisition and processiong
Read data  
```{r Activity}
library(dplyr)
library(ggplot2)
act<-read.csv("activity.csv")
# Number of days
numd<-unique(act$date)



```
## Mean total number of steps taken per day

- The total number of steps taken per day in variable actd is calculated.
- A histogram of the total number of steps taken each day is made.  
-The mean and median of the total number of steps taken per day is calculated.  
```{r Histogram steps per day}

actd<-as.data.frame(matrix(numd,length(numd),2))
colnames(actd)<- c("date","steps")
# na.rm is TRUE to ignor NA's, otherwise NA is returned. 
actd[,2]<-as.integer(tapply(act$steps,act$date,sum,na.rm=TRUE))
qplot(steps,data=actd,binwidth=1000,na.rm=TRUE  )
#Total number of steps per day is the mean per day of actd. Again NA'S are ignored
meand<-mean(actd$steps,na.rm=TRUE)
meand
mediand<-median(actd$steps,na.rm=TRUE)
mediand
```


 - The mean total number of steps per day is `r round(meand,2)` and the median of the total of steps per day is `r mediand`
 
## Average daily activity pattern
- A time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis) is made. 
- The 5-minute interval, on average across all the days in the data set, which contains the maximum number of steps is identified  
```{r Steps during the day}

numIn<-(unique(act$interval))
numIn<-numIn-as.integer(numIn/100)*40
#numIn<-ymd("2012-10-02")+hours(as.integer(numIn/100))+
#             minutes(numIn-as.integer(numIn/100)*100)


actIn<-as.data.frame(matrix(numIn,length(numIn),2))
colnames(actIn)<- c("Interval","steps")
actIn[,2]<-tapply(act$steps,act$interval ,mean,na.rm=TRUE)

qplot(Interval,steps,data=actIn,geom=c("line"))

```
   
 - The mean steps per day and
 - The 5-minute interval, on average across all the days in the data set, contains the maximum number of steps are calculated.   
```{r }
meanIn<-mean(actIn$steps)
Max<-actIn[which(actIn$steps==max(actIn$steps),arr.ind=TRUE),]
meanIn
actIn[as.numeric(row.names (Max)),]

```
The maximum is in the interval `r as.numeric(row.names (Max))` with a value of `r round(actIn[as.numeric(row.names (Max)),2],2)` steps

## Imputing missing values
The presence of missing days may introduce bias into some calculations or summaries of the data.

Calculate and report the total number of missing values in the data set (i.e. the total number of rows with N As)
```{r }
missteps<-which(is.na(act$steps),arr.ind=TRUE)
length(missteps)

```
The total number of rows with N As is `r length(missteps)`

The mean for that 5-minute interval is the strategy for filling in all of the missing values in the data set.
A new data set that is equal to the original data set but with the missing data filled in is created.
```{r }
# The na's are filled with the mean of all days for each interval 
act2<-act
for(i in missteps){
  nuin<-act2[i,3]
  act2[i,1]<-
    actIn[1+(nuin-as.integer(nuin/100)*40)/5,2]}
# Add one to consider  first intervale starts in 0 min.
# 40 is to consider 60 min per hour and 
# 5 is intervale duration.
anyNA(act2) # To check if there is any NA

```

A histogram of the total number of steps taken each day is made and the mean and median total number of steps taken per day are calculated and reported. 
```{r Histogram NAs filled}
actd2<-as.data.frame(matrix(numd,length(numd),2))
colnames(actd2)<- c("date","steps")
actd2[,2]<-as.integer(tapply(act2$steps,act2$date,sum))

qplot(steps,data=actd2,binwidth=1000  )
meand2<-round(mean(actd2$steps),2)
meand2
mediand2<-median(actd2$steps)
mediand2
```


Mean and median,`r round(meand2,2)`, `r mediand2`, do differ from the estimates from the first part of the assignment `r round(meand,2)`, `r mediand`   
The impact of imputing missing data on the estimates of the total daily number of steps is an increment of values. Since we are assuming a positive value instead of zero .


## Activity patterns between weekdays and weekends

The data set with the filled-in missing values is used for this part. A new factor variable in the data set is created with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r }
library (dplyr)
## start with all days aas weekdays and then change only the weekends
act2<-mutate(act2, tyday = "weekday")

weeke<-which(weekdays(as.Date(act2$date))=="sábado"|
               weekdays(as.Date(act2$date))=="domingo")
act2[weeke,4]<-"weekend"
act2$tyday<-as.factor(act2$tyday)
```
Panel plot is made containing a time series plot  of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r }
act2w<-data.frame(numIn,length(numIn),3)
names(act2w)<-c("Interval", "steps", "tyday")
act2wd<-data.frame(numIn,length(numIn),3)
names(act2wd)<-c("Interval", "steps", "tyday")

act3<-split(act2,act2$tyday)
act2w$steps<-tapply(act3$weekday$steps,act3$weekday$interval,mean)
act2wd$steps<-tapply(act3$weekend$steps,act3$weekend$interval,mean)

act2w$tyday<-"weekday"
act2wd$tyday<-"weekend"
act2w<-rbind(act2w,act2wd)

qplot(Interval,steps,data=act2w,facets=tyday~.,geom=c("line"))

```

## Conclusions

  - N As could bias results if they are ignored.
  - Considering and imputing of missing values with the average for each interval produces an increment on mean and median  with respect to ignore ring the N As.
  - there is a different activity pattern for weekdays and weekends. although the pattern is similar but intensity of activity is higher in weekdays. 
  





